---
output:
  html_document:
    keep_md: yes
---

###Getting and Cleaning Data - Project Writeup

####Introduction

We start out with seven files that must be combined to form one large flat file.  
Roughly speaking, there are:  
1. Three files for a training data set  
2. Three files for a test data set  
3. One file with the column headers for the previous two data sets  

These files are to be combined as shown in the figure in this forum post:
<a href="https://class.coursera.org/getdata-009/forum/thread?thread_id=58#comment-369">David's Project FAQ Diagram</a>

The data concerns internal sensors from a Samsung Galaxy II. 30 subjects each affix the device to their waist and then perform six activities:  
1. Walking  
2. Walking Upstairs  
3. Walking Downstairs  
4. Sitting  
5. Standing  
6. Laying  

More technical details are available at the home of the data set: <a href="http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones">UCI Machine Learning Repository</a>.  

After combining all the columns, we'll make a new data set containing only the columns containing a mean or a standard deviation (based on the column name).

Finally, we'll find the mean of each mean or st.dev. column by activity and subject. The final result should have 180 rows (30 subjects * 6 activities each) and many columns.


####Technical Analysis
The R code can be summarized neatly in four steps:  
1. Import all the necessary files  
2. Combine the seven files  
3. Filter on the mean and sd.dev columns  
4. Calculate the mean for each mean and sd.dev column

**Step 0 Set up knitr options**
```{r}
library(knitr)
opts_chunk$set(echo=TRUE,results="show",cache=TRUE,warning=FALSE)
```
Note that I am assuming all data files are in the same directory as the source file (this document), and knitr will make the working directory equal to the source file directory.

**Step 1 Import all the files for the testing sets**  
*1.1 Import subject_test.txt*  
```{r}
subject_test <- read.table("subject_test.txt",col.names="ID")
```
*1.2 Import X_test.txt*  
```{r}
xtest <- read.table("X_test.txt",header=F,nrows=2947,quote="e")
```
*1.3 Import y_test.txt*  
```{r}
ytest <- read.table("y_test.txt",header=F,nrows=2947)
```
*1.4. Import subject_train.txt*  
```{r}
subject_train <- read.table("subject_train.txt",col.names="ID")
```
*1.5. Import X_train.txt*  
```{r}
xtrain <- read.table("X_train.txt",header=F,nrows=7352,quote="e")
```
*1.6c. Import y_train.txt*  
```{r}
ytrain <- read.table("y_train.txt",header=F,nrows=7352)
```
*1.7. Read features.txt (This is the column names)*  
```{r}
library(data.table)
features <- fread("features.txt")
```

*1.8. Read activity_labels*  
Note: This step isn't technically required, but one goal is to provide understandable labels for each data element, and the six numeric activity codes are matched to words in this file.  
```{r}
activity_labels <- fread("activity_labels.txt")
setnames(activity_labels,old=c("V1","V2"),new=c("ID","activity_label"))
```

**Step 2 Put the Files Together**  
*2.1 Combine the subject, X, and y testing files with cbind*
```{r}
testFile <- cbind(subject_test,xtest,ytest)
```
*2.2 Do the same with the training files*
```{r}
trainFile <- cbind(subject_train,xtrain,ytrain)
```
*2.3 Combine those into one file with rbind*  
(Note: This step generates a huge annoying warning. I don't know why, but it's 100% about columns that won't be kept in Step 3, so I don't care. =D)  
```{r}
finalFile <- rbind(trainFile,testFile)
```
*2.4 Rename the columns using the column names in the second row of the features file and manually name the subject (1st column) and activity (last column)*  
```{r}
names(finalFile) <- c("subject",features$V2,"activity")
```
*2.5. Let's also merge in the activity labels*  
(It's actually a waste of memory to do that now, but for this assignment it seems to fit with readability here..)  
```{r}
finalFile <- merge(finalFile,activity_labels,by.x="activity",by.y="ID")
```


**Step 3 Extract only the mean and standard deviation columns**  

*3.1 Make a list of the column names (excluding subject, activity, and activity_label)*
```{r}
myNames <- names(finalFile[1,3:563])
```
*3.2 Get the mean and sd.dev columns*  
Here I'm using regular expressions to tell me which columns contain the literal string "mean" and "std", then make a T/F vector where either is true. Those will be the data columns I want to keep.  
```{r}
myMeanCols <- regexpr("mean",myNames,ignore.case=T) != -1
mySDCols <- regexpr("std",myNames,ignore.case=T) != -1
myCols <- myMeanCols | mySDCols
```

*3.3 Extract the desired columns*  
Note I also manually keep activity, activity_label, and subject. Only the data columns were checked for in 3.2  
Note 2: I'm using setnames() throughout this document because 'colnames(filterFile) <- whatever' was throwing warnings uring the change.  
```{r}
filterFile <- finalFile[,c(T,T,myCols,T)]
n <- names(filterFile)
n2 <- make.names(n,unique=T)
setnames(filterFile,old=n,new=n2)
```
**Step 4 Summarize**  
I'll use aggregate to average means and sd's by activity and subject, then rename the columns with the existing names plus 'meanOf_'.  
e.g. 'tBodyAcc.mean...X' becomes 'meanOf_tBodyAcc.mean...X' since it is the mean of the raw data values.  
```{r}
aggFile <- aggregate(filterFile[,3:88],
                     by=list(activity=filterFile$activity,
                             activity_label=filterFile$activity_label,
                             subject=filterFile$subject),
                     mean)
n <- names(aggFile)
n2 <- sapply(n[4:89],function(elt) paste("meanOf_",elt,sep=""))
setnames(aggFile,old=n,new=c(n[1:3],n2))

```
The resulting file has 180 rows: 6 activities for each of 30 subjects.  
Every column is named appropriately, and all means and standard deviation columns in the original data are represented with their means by subject and activity.  
Mission complete. [sunglasses emoji]  